The zoon R package for reproducible and shareable species distribution modelling

### 2016-12-20 ms.Rmd
remove codeface from 'zoon' in text

---
title: "The `zoon` R package for reproducible and shareable species distribution modelling"
author: "Nick Golding, Tom August, Tim C.D. Lucas, Emiel Van Loon & Greg McInerny"
bibliography: zoon_app.bib
csl: mee.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex_preamble.tex
---

@@

# The `zoon` R package

@@

The `zoon` R package encodes SDM analyses as a simple workflow of five key steps, with each step encoded as one or more software modules. Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis. Figure 1 illustrates the structure of a `zoon` workflow comprising the five module types and illustrates how multiple modules may be combined at each stage. The module types and module combination options are described in more detail below.

@@

```{r flow_diagram, echo = FALSE, out.width = "375px", fig.cap = "The modular SDM structure encoded by a zoon workflow. A) Flow diagram representing the five module types and intermediate objects: df - an occurrence dataframe; ras - a RasterStack object; mod - a ZoonModel object; ? - any user-defined inputs or outputs. B) The flow diagram implied by chaining modules of the same type. C) The flow diagram implied by listing modules of the same type. Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the `zoon` vignette 'Building a module'."}
knitr::include_graphics("figs/diagrams.png")
```

@@

@@

Many SDM analyses require running similar workflow in parallel. 
In applied settings this may be running the same analysis for multiple species.
In methodological work, it is important to compare new methods or models to benchmarks.
To divide a `zoon` workflow into multiple parallel analyses the `list` function is used.
`list` can be used to split the analysis at any point: listing multiple occurrence modules together will run the workflow on different occurrence datasets while listing multiple model modules will fit different models to a shared, identical dataset but then apply output modules to each model separately. In cases where modules containing randomness need to be run many times - such as running a workflow with hundreds of simulated datasets - `replicate` can be used. This simply creates a `list` of the specified module, repeated a given number of times.

@@

### Inspecting, sharing and extending a `zoon` workflow

@@

`zoon` workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
A `zoon` workflow returns an object of class `zoonWorkflow`.
This object contains all the information needed to reproduce the analysis.
This includes the original call, a record of the packages and modules used and their versions when the analysis was run as well as all the data and fitted models.

@@

`zoon` provides functionality to quickly update an existing workflow object (even one created by someone else) by switching out modules.

@@

As long as a valid module function is defined in an R session, it can be used in workflows in combination with modules downloaded from the `zoon` module repository.
Whilst this makes it easy to develop modules locally, the strength of `zoon` comes from the ability to upload modules to the online repository so that others can access them.

To do this, it first is necessary to provide additional metadata including the module tile and description, module type, author name and definitions for any non-mandatory arguments.
`zoon` provides the function `BuildModule()` to facilitate entering this metadata and checking that it matches the module.

The `zoon` vignette *Building modules* provides a detailed tutorial for building modules of each type.

@@

`zoon` provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare. `ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object, along with a metadata text file, to the user's figshare profile for others to download, inspect and modify.

@@

We demonstrate the `zoon` R package by recreating two SDM analyses from published research articles as `zoon` workflows and showing how these workflow objects can easily be extended with new modules.
Workflow objects created by these analyses can be accessed at [`figshare.com/articles/zoon_applications_paper_workflows`](http://figshare.com/articles/zoon_applications_paper_workflows).
We encourage readers to download, interrogate and alter these workflows for themselves.
Full code and metadata for all of the modules used in the examples below can be found at [`github.com/zoonproject/modules/R`](https://github.com/zoonproject/modules/R)

@@

Such a model can be quickly and easily re-constructed as a `zoon` workflow using modules available in the `zoon` module repository.
@Feng2015 used a combination of occurrence data from GBIF, and additional occurrence data manually collected from the published literature.
Unfortunately the latter data have not been made publically available, so here we use only data from GBIF.
If the additional data had been made available it would be straightforward to incorporate them, for example using the `LocalOccurrenceData` module.

@@

@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data, by 'thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the `zoon` data into the spThin format, uses this package to apply the algorithm, and converts data back into `zoon`'s expected format again:

@@

To convert this code into a `zoon` module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation. The `zoon` function `BuildModule()` helps with this step, and can also run checks to make sure we got everything right:

@@

This module can now be shared so that others can use it in their `zoon` workflows. The best way of sharing it is via the `zoon` modules repository. Modules can uploaded there, via the module submission system at [URL TBD].

@@

In this example we illustrate how `zoon` can help contribute to discussions such as these by enabling rapid evaluation of a new method against a variety of datasets and test procedures.

@@

All of these have now been uploaded to the `zoon` modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.

@@

However, “scientific software is developed either by domain scientists who have little training in user interface (UI) or experience (UX) design, or by software engineers who build broad systems that do not suit the niche needs or practices of domain scientists” [@Thomer2016].
Then, there is a need for environments that can address niche needs (e.g. of a modelling discipline such as SDM) and that address usability issues for diverse outputs produced by diverse scientists.
The SDM community is large enough to warrant a discipline specific platform such as `zoon`.
The Zoon Project offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.
It may be that zoon can also be used for other modelling activities that have similar workflows, or use similar data and processing steps.

During the development of `zoon` we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
This is by no means a full UX approach [@Pavelin2012], but it has directly informed the design. 
Continued user engagement will be vital to developing a critical mass of `zoon` modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
An important step would be the publication of a large model inter-comparison within the `zoon` environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
In order to facilitate a user –friendly approach to distributed working, future work will include the development of an online interface for accessing a repository of contributed modules and modifying workflows.
This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.
In sum, we hope that the `zoon` R package represents a step towards a more reproducible ecosystem of SDM software.

To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with `zoon` – including  guides on creating workflows and modules, as well as full technical details for module developers.

>>>>

---
title: "The zoon R package for reproducible and shareable species distribution modelling"
author: "Nick Golding, Tom August, Tim C.D. Lucas, Emiel Van Loon & Greg McInerny"
bibliography: zoon_app.bib
csl: mee.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex_preamble.tex
---

@@

# The zoon R package

@@

The zoon R package encodes SDM analyses as a simple workflow of five key steps, with each step encoded as one or more software modules. Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis. Figure 1 illustrates the structure of a zoon workflow comprising the five module types and illustrates how multiple modules may be combined at each stage. The module types and module combination options are described in more detail below.

@@

```{r flow_diagram, echo = FALSE, out.width = "375px", fig.cap = "The modular SDM structure encoded by a zoon workflow. A) Flow diagram representing the five module types and intermediate objects: df - an occurrence dataframe; ras - a RasterStack object; mod - a ZoonModel object; ? - any user-defined inputs or outputs. B) The flow diagram implied by chaining modules of the same type. C) The flow diagram implied by listing modules of the same type. Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the zoon vignette 'Building a module'."}
knitr::include_graphics("figs/diagrams.png")
```

@@

<!— remove this, it's covered in the figure —>

@@

Many SDM analyses require running similar workflow in parallel. 
In applied settings this may be running the same analysis for multiple species.
In methodological work, it is important to compare new methods or models to benchmarks.
To divide a zoon workflow into multiple parallel analyses the `list` function is used.
`list` can be used to split the analysis at any point: listing multiple occurrence modules together will run the workflow on different occurrence datasets while listing multiple model modules will fit different models to a shared, identical dataset but then apply output modules to each model separately. In cases where modules containing randomness need to be run many times - such as running a workflow with hundreds of simulated datasets - `replicate` can be used. This simply creates a `list` of the specified module, repeated a given number of times.

@@

### Inspecting, sharing and extending a zoon workflow

@@

zoon workflows are re-runnable records of the data, code and results of an entire SDM analysis and can be easily reproduced, scrutinized and extended by the whole research community.
A zoon workflow returns an object of class `zoonWorkflow`.
This object contains all the information needed to reproduce the analysis.
This includes the original call, a record of the packages and modules used and their versions when the analysis was run as well as all the data and fitted models.

@@

zoon provides functionality to quickly update an existing workflow object (even one created by someone else) by switching out modules.

@@

As long as a valid module function is defined in an R session, it can be used in workflows in combination with modules downloaded from the zoon module repository.
Whilst this makes it easy to develop modules locally, the strength of zoon comes from the ability to upload modules to the online repository so that others can access them.

To do this, it first is necessary to provide additional metadata including the module tile and description, module type, author name and definitions for any non-mandatory arguments.
zoon provides the function `BuildModule()` to facilitate entering this metadata and checking that it matches the module.

The zoon vignette *Building modules* provides a detailed tutorial for building modules of each type.

@@

zoon provides the `ZoonFigshare` function to facilitate sharing a completed workflow object from within R, via the free web platform figshare. `ZoonFigshare` takes a workflow object and some minimal metadata and uploads the workflow as an RData object, along with a metadata text file, to the user's figshare profile for others to download, inspect and modify.

@@

We demonstrate the zoon R package by recreating two SDM analyses from published research articles as zoon workflows and showing how these workflow objects can easily be extended with new modules.
Workflow objects created by these analyses can be accessed at [`figshare.com/articles/zoon_applications_paper_workflows`](http://figshare.com/articles/zoon_applications_paper_workflows).
We encourage readers to download, interrogate and alter these workflows for themselves.
Full code and metadata for all of the modules used in the examples below can be found at [`github.com/zoonproject/modules/R`](https://github.com/zoonproject/modules/R)

@@

Such a model can be quickly and easily re-constructed as a zoon workflow using modules available in the zoon module repository.
@Feng2015 used a combination of occurrence data from GBIF, and additional occurrence data manually collected from the published literature.
Unfortunately the latter data have not been made publically available, so here we use only data from GBIF.
If the additional data had been made available it would be straightforward to incorporate them, for example using the `LocalOccurrenceData` module.

@@

@Aiello-Lammens2015 proposed an approach for dealing with spatial sampling bias in presence-only data, by 'thinning' the presence data, and provide an R package `spThin` to implement their procedure [@spThin].
We can incorporate this approach in a workflow, by defining a simple *process* module that adapts the zoon data into the spThin format, uses this package to apply the algorithm, and converts data back into zoon's expected format again:

@@

To convert this code into a zoon module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation. The zoon function `BuildModule()` helps with this step, and can also run checks to make sure we got everything right:

@@

This module can now be shared so that others can use it in their zoon workflows. The best way of sharing it is via the zoon modules repository. Modules can uploaded there, via the module submission system at [URL TBD].

@@

In this example we illustrate how zoon can help contribute to discussions such as these by enabling rapid evaluation of a new method against a variety of datasets and test procedures.

@@

All of these have now been uploaded to the zoon modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.

@@

However, “scientific software is developed either by domain scientists who have little training in user interface (UI) or experience (UX) design, or by software engineers who build broad systems that do not suit the niche needs or practices of domain scientists” [@Thomer2016].
Then, there is a need for environments that can address niche needs (e.g. of a modelling discipline such as SDM) and that address usability issues for diverse outputs produced by diverse scientists.
The SDM community is large enough to warrant a discipline specific platform such as zoon.
The Zoon Project offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.
It may be that zoon can also be used for other modelling activities that have similar workflows, or use similar data and processing steps.

During the development of zoon we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
This is by no means a full UX approach [@Pavelin2012], but it has directly informed the design. 
Continued user engagement will be vital to developing a critical mass of zoon modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by @Elith2006 or more recently in @Beaumont2016.
At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
In order to facilitate a user –friendly approach to distributed working, future work will include the development of an online interface for accessing a repository of contributed modules and modifying workflows.
This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.
In sum, we hope that the zoon R package represents a step towards a more reproducible ecosystem of SDM software.

To assist users a number of tutorials are provided on the GitHub repository and as vignettes distributed with zoon – including  guides on creating workflows and modules, as well as full technical details for module developers.
